Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-14 article.tex
stack order:  code and write-up; lame conclusion

\section{Monad Transformer Stacks}

1. Exact vs. relative layer order; 
2. one instance per stack, even when more than one is applicable --
how to specify which one?
3. an inappropriate instance higher-up in the stack can mask a 
desired instance lower down in the stack -- that's why we had to
create a replacement for the Alternative type class and provide
some new instances for it -- to be able to use the right instances

\section{Transformer instance semantics}

Pass-through vs. deal-with, examples of each

\section{Conclusion}
Furthur applications of monad transformers to parsing:  junk logging, partial results, 
 non-deterministic parsing using the ListT or LogicT or just list data type

>>>>

\section{Stack Order}
Not only are the contents of a transformer stack important, but the order of the
layers as well.  Let's revisit our first woof parser -- what if our stack had
instead been Maybe/State instead of State/Maybe?
\begin{verbatim}
State/Maybe:   s -> Maybe (a, s)

Maybe/State:   s -> (Maybe a, s)
\end{verbatim}

given:
\begin{verbatim}
import Control.Monad.Trans.Maybe (MaybeT(..))
import Data.Functor.Identity     (Identity(..))

type Parser' t a = MaybeT (StateT [t] Identity) a

runParser' :: Parser' t a -> [t] -> (Maybe a, [t])
runParser' p xs = runIdentity $ runStateT (runMaybeT p) xs
\end{verbatim}
We can compare it to the earlier parser:
\begin{verbatim}
*Main> runParser' (literal 'a') "aqrs"
(Just 'a',"qrs")

*Main> runParser' (literal 'a') "qrs"
(Nothing,"rs")
\end{verbatim}
Whereas:
\begin{verbatim}
*Main> runParser (literal 'a') "qrs"
Nothing

*Main> runParser (literal 'a') "aqrs"
Just ('a',"qrs")
\end{verbatim}
The difference is that Maybe/State always returns a modified state, regardless
of the success of the computation, whereas State/Maybe only returns a modified
state if the computation is successful.  This means that backtracking doesn't
work right with Maybe/State -- tokens are consumed even when the parsers
don't match, as the example shows.

On the other hand, the order of Error and Maybe relative to each other isn't 
important.  Try it for yourself!

Another important point to note is that our parsers didn't actually specify
which order -- Maybe/State or State/Maybe -- they prefer, and in fact they will 
work with either order.  Since our parsers don't specify the stack order, the 
semantics are ultimately determined by the actual stack we use.

\section{Conclusion}
Monad transformers provide an elegant solution to error reporting in parser
combinators.  Better yet, they are extensible -- we could go back and add more
features to our parsers using additional transformers, perhaps extending our
parser to save partial results, so that if there's an error, it reports the
progress that it had made before the error.  We could do that using a Writer
transformer.  Or we check to make sure that all variables are in scope when
they're used -- using a Reader transformer.  We could even capture whitespace
and comment tokens, instead of just throwing them away, in case they contained
valuable information.  And the best part of it is, most of the parsers we already
wrote would continue to work fine without needing any modifications.  We'd just
have to set up our transformer stack and modify the relevant parsers!
